Comparison of large-scale citizen science data and long-term study data for phenology modeling

### 2018-09-19 manuscript/Reviewer_Reply.md

We would like to thank the editor and two reviewers for the  productive feedback on our manuscript. We have addressed comments related to the sensitivity of the 30-day cutoff in using the USA-NPN data, and included general recommendations for which models performed best in different contexts. More detailed responses to individual comments are included below.

@@

**General comment: I appreciate the authors' efforts to ensure transparency and repeatability of the analysis.**

Thank you.

**l. 20: not all the models are process-based; revise?**

We changed this to "statistical and processed based models".

@@

**l. 127: phenophase (singular)**

Fixed.

**ll. 210-211. I don't understand this. Why do some seemingly strong relationships (e.g., "naive" panel of Fig. 2; also some panels of Fig. 3) yield negative R-squared values? This probably just reflects my ignorance... but it's likely that if I'm mystified, other readers will be, too! Please explain more clearly how the <0 R-squareds arise.**

These values are coefficients of determination which compare the predictions from the model to the observed values. If there is a strong bias in the predictions this can lead to negative values even if the observations and predictions are correlated. Specifically, in cases where choosing the mean value of the observations provides a better prediction than the model these numbers will be negative. While this is not possible in simple regression without held out data (because the model by definition does at least as well as the mean) it is possible for other kinds of models and in cases where the model is not fit to the data it is evaluated on. We have attempted to clarify this in the text.

@@

# Reviewer 1 Comments to Author:

### Comments to the Author
**Comparison of large-scale citizen science data and long-term study data for phenology modeling**

**This is a very practical study and the evaluation of the approach is clear and pretty straightforward to follow. The authors might improve readability a bit by either adding subheads into the Methods & Results or repeatedly using the same terms to refer to the various steps (eg, cross-validation).**

After some discussion, we decided not to add subheadings as both sections are relatively short, and each have a single paragraph dedicated to each analysis step. Instead we have made the Analysis introduction paragraph lay out the three primary steps, and also adjusted the wording to be more consistent.

@@

**The most critical thing that I see that’s lacking is that there is no consideration of how decisions made regarding which NPN data to include in the analysis might be impacting model performance. The authors allowed observations with up to 30d between the last reported “no” and the first reported “yes” and chose the midpoint. This could still lead to up to 15d error in the assumed date of “yes” and the true date of “yes”; this variance may in many cases be larger than the signal in the model. Though actually addressing this sensitivity is probably beyond the scope of the analysis, the potential and implications for this error should be discussed explicitly in the paper.**

We tested the sensitivity of this cutoff time by repeating the analysis with a 15 day cutoff. This change did not influence our results. Three species (flowers and budburst for Vaccinium parvifolium and  budburst for Acer circinatum) were dropped from the 15 day analysis due to inadequate sample size. The overall sample size for the 15 day analysis was decreased by 10% (619 observations) and by 1-35% for individual species. We have added notes on this sensitivity analysis to the methods and results as well as reporting the results in new supplementary material.

@@

**The authors lay out two goals for this effort – to determine “inferences about biological processes driving phenology” and to develop predictive phenology models. It may be just me…using the term “inference” in this way confused me.**

This is a common usuage in the data science community and we are unclear how to better communicate this distinction between making predictions and understanding processes. To avoid confusion we have added a specific example to describe our usage of inference in the last introduction paragraph.

@@

**Regarding the predictions that can result from these different models…this is of strong interest to many stakeholder realms. It would be fantastic to see a bit more attention given in the discussion to recommendations for implementation. The authors talk about the tradeoffs between more complex formulations and simpler models and – generally – reasons these differences may come about. It would be great to also see a bit of more practical discussion along the lines of, “if you can tolerate error of 7-14d, then model X may meet your needs, and it’s fine to build the models with either NPN or LTER data. However, if your needs are Y, then you’d be better off doing blah, blah…”… would this be possible?**

We have added additional text to the Discussion describing the best models from our analysis and pointed readers to the supplemental images for specific details of model performance. We have also updated supplemental images S1 and S2 (now S5 & S6) to better show model performance in the cross-dataset comparison, and also added an additional table (Table S2) which gives the overall model performance across all species and phenophases.

@@

**In several places “data” is treated as singular… data are plural**

Fixed.

**L67: the citizen science program run by the USA-NPN that yields the data housed in the National Phenology Database is Nature’s Notebook**
**Thank you for acknowledging the Nature’s Notebook contributors for the data!**

You're welcome. We're happy to credit all data providers, as this study would not have been possible without them. We have adjusted this line of text to reflect that Nature's Notebook is the citizen science program run by USA-NPN.

@@

**Please see the USA-NPN data attribution policy for recommendations for how to properly cite the dataset: https://www.usanpn.org/terms#DataAttribution**

We have added the correct citation for the dataset (USA National Phenology Network, 2017)

### Abstract
**be sure to indicate that the study is focused on the U.S.**

This is now made clear by specifying that it's the USA National Phenology Network.

@@

**I don’t think I saw “RMSEA – RMSEB” and “RMSEC-RMSED” referenced anywhere else in the paper**

We removed these specific equations and replaced them with a better description of this analysis.

**L196: what does “scenario A” refer to here?**

This referred to the RMSE values derived from LTER Model predictions for held out LTER data. We rewrote this paragraph and changed the labels on Figure 4 to make this analysis more clear.

@@

**While I do think that this manuscript should be ultimately acceptable for publication at Ecology, there are a few major/minor comments and suggestions for which I would like the authors to consider.**

Major Comments and Suggestions:

**First, for NPN species with larger pools/spatial extents of observations, I wonder if would be useful to calibrate the models using subsets somehow stratified by climate and/or latitude. For example, train the models using only warm region (or southern) red maple and test on cold region maples (or vice versa). Perhaps this could further address the issue of model identifiability (as discussed in Lines  259-277). While the oft-mentioned Basler (2016) study certainly had nice spatiotemporal coverage, it did not have nearly the latitudinal/climate variability that NPN affords.**

We agree that this approach is a valuable next step, but ultimately decided that doing this properly would require a follow up paper rather than an additional analysis in this one. That said, we are doubtful that it would address the issue of identifiability. The Harvard and Hubbard Brook LTER share 3 species in common and are relatively close (~150km), yet their parameter estimates varied widely for most models (see images S4-S5 in the original text, now S8-S9 in the revised text).

@@

**Second, given the length of time series of the LTER datasets, I think it would be useful to further examine the performance of models based on other criteria such as the amount of interannual variability explained (correlation or R-square) or bias during anomalously warm or cold years. Perhaps the latter might be outside of the overall realm of the paper, but I think it would be informative to know which models capture this type of information which matters in a global change context.**

See the note to Reviewer 1 about general model recommendations. We have updated the Supplemental images S1 and S2 to included Pearson's correlation in addition to the RMSE for all dataset inter comparisons. We decided that more expansive analyses were beyond the scope of this paper, but are hopeful that the open and reproducible nature of the code and data for this project will allow others (or our future selves) to conduct the important follow up analyses suggested by the reviewer.

### 2018-09-19 manuscript/main.tex
final version for 1st resubmission

Data from citizen science projects are becoming increasingly important for ecological research \citep{kelling2009, dickinson2010, tulloch2013}. Because these data are often collected by large numbers of volunteers, it is possible to gather data at much larger scales than with individual research teams. A relatively new citizen science project started in 2009, The USA National Phenology Network (USA-NPN), collects phenology observations from volunteers throughout the United States and makes the data openly available \citep{npncitation}. Data from this project have already been used to study variation in oak phenology at a continental scale \citep{gerst2017}, develop large-scale community phenology models \citep{melaas2016}, and forecast long-term phenology trends \citep{jeong2013}. Large-scale datasets from China and Europe have already contributed considerably to phenological research \citep{xu2013, olsson2014, basler2016, zhang2017}, and the USA-NPN dataset has the potential to meet these needs for North American plant species and communities. However, the features that allow citizen science projects to collect data at large scales can also introduce spatial biases toward cities and easily-accessible areas, and variation in sampling effort and observer skill \citep{dickinson2010}. With thousands of participants, the potential for variation among observers in their determination of species identification and dating of phenological events is high. While volunteers have been shown to be accurate at distinguishing different leaf and flower stages for plants \citep{fuccillo2015} and can have high agreement on abundance estimates \citep{feldman2018}, contributions to USA-NPN are sometimes made sporadically across seasons, years, and locations. This means that the quantity and quality of data at a specific site will typically be more variable for citizen science efforts than for intensive, long-term studies.

In order to accurately model and forecast phenology, it is important to understand how the strengths and weaknesses of intensive local studies and large-scale citizen science projects influence both our inferences about biological processes driving phenology (eg. warming requirements for a specific plant) and our ability to predict future phenology events (eg. forecasting when flowering or leaf out occurs). Here, we fit a suite of plant phenology models for the budburst and first-flowering phenophases of 24 plant species to data from both the USA-NPN and a set of intensive long-term studies from the Long Term Ecological Research (LTER) network. We compare the resulting models based on both inference about models and parameters and predictions for unobserved events. We then use this comparison to assess the best methods for both local- and large-scale phenology modeling and to point the way forward for integrating large-scale and local-scale data to determine the best possible models across scales.

@@

In the LTER datasets observation metrics varied widely due to different protocols. To match the USA-NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods). As with the USA-NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophase combinations which had at least 30 total observations. After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the USA-NPN and LTER datasets to use in the analysis (Table 1 \& S1). Using a 15 day cutoff in the USA-NPN dataset resulted in 35 unique combinations with 23 species .

@@

It is common to fit multiple plant phenology models to find the one that best represents a specific species and phenophase \citep{chuine2013}. For each of the 38 species and phenophase combinations in the five datasets (USA-NPN and four LTER datasets), we fit eight phenology models (Table 2). The \textit{Naive model} uses the mean DOY from prior observations as the estimated DOY. The \textit{Linear model} uses a regression with the mean spring (Jan. 1 - March 31) temperature as the independent variable and DOY as the response variable. For the six remaining models, the general form is based on the idea that a phenological event will occur once sufficient thermal forcing units, $F^{*}$, accumulate from a particular start day of the year ($t_{1}$). Forcing units are a transformation of the daily mean temperature and are calculated differently for each model (Table 2). The start day can either be estimated or fixed. For the \textit{Growing Degree Day (GDD)} model, forcing units are the total degrees above the threshold $T_{base}$ \citep{reaumur1735, wang1960, hunter1992}. The \textit{Fixed GDD model} uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T_{base}$ = 0$^{\circ}$C). The \textit{Alternating model} has a variable number of required forcing units defined as a function of the total number of days below 0$^{\circ}$C since Jan. 1 (number of chill days - $NCD$). The \textit{Uniforc model} is like the GDD model but with the forcing units transformed via a sigmoid function \citep{chuine2000}.

@@

As described above, we fit two sets of models for each species and phenophase: one set of models parameterized using only USA-NPN data, and one set parameterized using only LTER data (with the exception of the \textit{M1} and \textit{MSB} models, see above). To compare the inferences about process made by the two datasets, we compared the distribution of each parameter between LTER and USA-NPN derived models for each species and phenophase combination. Using the mean value of each bootstrapped parameter, we also calculated the coefficient of determination ($R^2$) between LTER and USA-NPN derived models among the 38 species-phenophases. In three cases where a species phenophase combination occurred in two LTER sites (Budburst for \textit{Acer saccharum}, \textit{Betula alleghaniensis}, and \textit{Fagus grandifolia} in the Harvard and Hubbard Brook datasets) they were compared separately to the USA-NPN data.

Models with different parameter values, and even entirely different structures, can produce similar estimates for the date of phenological events \citep{basler2016}. Therefore, to compare the predictions and potential forecasts for models fit to the different datasets, we compared the estimated DOY predicted by the LTER and USA-NPN derived models for all held out observations. For each of the eight models, we calculated the coefficient of determination ($R^2$) between LTER and USA-NPN derived estimates for estimates made at the four LTER sites and across all USA-NPN sites.

We also directly evaluated model performance using the four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) USA-NPN derived models predicting LTER observations, C) LTER-derived models predicting USA-NPN observations, and D) USA-NPN derived models predicting USA-NPN observations. Using the RMSE values from held out observations, we  compared the  performance of LTER and USA-NPN derived models on different data types in two different ways. First, we focused on local-scale prediction by calculating the difference in the RMSE of LTER and USA-NPN derived models solely with LTER observations. Secondly, to focus on large-scale prediction we calculated the difference in RMSE using solely USA-NPN data. These differences were calculated for each of the model types and 38 species-phenophase combinations. Negative values indicate that LTER-derived models perform better, while positive values indicate that the USA-NPN derived model performed better. We used a t-test to test the difference from zero in these values. In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison within each site, not across sites, to focus on local scale prediction when LTER data are available. Absolute RMSE values are provided in the supplement (Fig. S5-S7).

@@

Throughout the analysis no significant difference was found between a 30 day or 15 day threshold between the first 'yes' and most recent 'no' observation in the USA-NPN dataset. Results presented here reflect the 30 day cutoff, see the supplemtary figures S2-S4 for matching figures using a 15 day cutoff. The best matches between parameter estimates based on USA-NPN and LTER data were the \textit{Fixed GDD model} ($R^2$ = 0.49) and the \textit{Linear model} ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$). The parameters for all other models had $R^2$ values \textless 0 indicating that the relationship was worse than no relationship between the parameters (but with matching mean parameter values across the two sets of models) (Fig. 2). The \textit{Naive model} showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTER datasets being mostly in the northern United States compared to the site locations of the USA-NPN dataset (Fig. 2). The large outlier for the \textit{Fixed GDD model} is \textit{Larrea tridentata}; this species' flower phenology is largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}. While the \textit{Fixed GDD} and \textit{Linear} models showed reasonable correspondence between parameter estimates, all parameters for individual species and phenophase combinations had different distributions between USA-NPN and LTER-derived models (Fig. S10-S11).

@@

When comparing the prediction accuracy on held-out data, USA-NPN derived models made more accurate predictions for held-out USA-NPN observations, and LTER-derived models performed better on held-out LTER observations ( all \textit{p} < 0.001, Fig. 4). The \textit{Naive} and \textit{Linear} models had the largest differences between the two model sets, while the \textit{Fixed GDD model} had relatively similar errors when evaluated on both USA-NPN and LTER held-out observations. Although the \textit{Fixed GDD model} had the highest agreement in accuracy between USA-NPN and LTER-derived models, it was not the best performing model overall. The \textit{GDD} and \textit{Uniforc} models commonly made the best predictions, having the lowest RMSE in 23\% and 40\% of cases among USA-NPN derived models, and 42\% and 32\% of cases among LTER-derived models, respectively (Fig. S5 \& S6).

@@

In this study, parameter estimates differed widely within the same phenology model when fit to the two different types of data, except for the simplest process-oriented model: the Fixed GDD (Fig. 2). These differences may be caused by a variety of factors that have different implications for interpreting process-oriented models and their parameters. First, the differences could result from limitations in the sampling of the USA-NPN dataset, such as irregular sampling of the same location within or between seasons, leading to less accurate parameter estimates. If this is the case, it would suggest that using LTER data is ideal for making inferences about plant physiology, and that focusing on the Fixed GDD model is best for making inferences when USA-NPN data are all that is available. Second, spatial variation (eg. from local adaptation, acclimation, microclimates, or plant age)  in phenology requirements and drivers could contribute to these differences \citep{diez2012, zhang2017}. Model built using USA-NPN data integrate over that spatial variation, while models built using LTER data only estimate the phenological requirements for a specific site. In this case, USA-NPN data would provide a better estimate of the general phenological requirements of a species, but LTER data would provide a more accurate understanding for a single site. The best solution to this issue would be the development of models that accurately incorporate spatial variation, such as including genetic variation between different populations \citep{chuine2017}, although localized models could also be generated when large-scale predictions are unnecessary. Third, these differences could result from issues with model identifiability: since different parameter values can yield nearly identical estimates of phenological events, parameter estimates can differ between datasets even when the underlying processes generating the data are the same. Information about which of these issues may be causing the differences between datasets can be explored using the analyses in the current study, as will be explained below.

Despite substantial differences in parameter estimates, LTER and USA-NPN derived models produced similar estimates for phenological events in most cases (Fig. 3). This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues. For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T_{base}$=0 and $t_{1}$=5, $F$=5, $T_{base}$=0 produce nearly identical estimates in many scenarios. This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters. In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both USA-NPN and LTER data (Fig. S10-S11). Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Fig. S8-S9). These results are consistent with research showing that models failed to estimate the starting day of warming accumulation solely from budbreak time-series, thus producing parameter estimates which were not biologically realistic \citep{chuine2016}. \cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting. Here, our simplest model, the \textit{Fixed GDD model} which uses only a warming component, had the highest correlation among parameters between LTER and USA-NPN datasets. In combination with the aforementioned studies, our results indicate that caution is warranted in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.

@@

In addition to exploring differences between phenology datasets our analysis provides guidance on which models to use when making predictions at a local scale using models built from large scale data, or vice versa. Among the eight models tested the Uniforc model performed the best overall in the cross dataset comparison in terms of pearson correlation and RMSE (Table S2). One exception was the Fixed GDD model had a slightly higher pearson value when using LTER derived models to make predictions for USA-NPN data. When the scale is the same between predictions and training data (ie. using LTER derived models to make predictions for LTER data) the Uniforc, GDD, and M1 models performed equally well, though the Uniforc had slightly better RMSE (Table S2). The best model for specific species and phenophases varied, but was commonly the Uniforc and GDD models (Figs. S5-S6).

In conclusion, our results suggest that both LTER and USA-NPN data provide valuable information on plant phenology. Models built using both data sources yield effective predictions for phenological events, but parameter estimates from the two data sources differ and models from each source best predict that data source's phenology events. The primary difference in the datasets is spatial scale, but due to trade-offs in data collection efforts, the larger scale USA-NPN data have shorter time-series, less site fidelity and other differences from the intensively collected LTER data (Table 3). These differences can be strengths or potential limitations. Observers sampling opportunistically allows the USA-NPN dataset to have a large spatial scale, but also leads to low site fidelity which limits the ability to measure long-term trends at local scales \citep{gerst2016}. Tracking long-term trends is the major strength of LTER data, but having a relatively small species pool limits their use in species-level predictive modeling. Due to these differences, the best data source for making predictions depends on the scale at which the predictions are being made. Identifying the most effective data sources for different types and scales of analysis is a useful first step, but the ultimate solution to working with diverse data types is to focus on integrating all types of data into analyses and forecasts \citep{hanks2018, melaas2016}. Our results suggest that methods that can learn from the intensive information available in LTER data in regions where they are available, and simultaneously use large-scale data to capture spatial variation in phenological requirements will help improve our ability to understand and predict phenology. Data integration efforts should also leverage data from remote sensing sources such as the PHENOCAM network or satellite imagery, which have both a large spatial extent and high temporal resolution \citep{peng2017,richardson2018a,richardson2018b}. Data integration provides the potential to use data from many sources to produce the best opportunity for accurate inference about, and forecasting of, the timing of biological events.

@@

This research was supported by the Gordon and Betty Moore Foundation’s Data-Driven Discovery Initiative through Grant GBMF4563 to E.P. White. We thank the developers and providers of the data that made this research possible including: the USA National Phenology Network and the many participants who contribute to its Nature’s Notebook program; the H.J. Andrews Experimental Forest research program, funded by the National Science Foundation's (NSF) LTER Program (DEB-1440409), US Forest Service Pacific Northwest Research Station, and Oregon State University; the Jornada Basin LTER project (NSF Grant DEB-1235828); the Hubbard Brook Experimental Forest, which is operated and maintained by the USDA Forest Service, Northern Research Station, Newtown Square, PA.; the Harvard Forest LTER; and the PRISM Climate Group at Oregon State University.

>>>>

Data from citizen science projects are becoming increasingly important for ecological research \citep{kelling2009, dickinson2010, tulloch2013}. Because these data are often collected by large numbers of volunteers, it is possible to gather data at much larger scales than with individual research teams. A relatively new citizen science project started in 2009, Nature's Notebook run by The USA National Phenology Network (USA-NPN), collects phenology observations from volunteers throughout the United States and makes the data openly available \citep{npncitation}. Data from this project have already been used to study variation in oak phenology at a continental scale \citep{gerst2017}, develop large-scale community phenology models \citep{melaas2016}, and forecast long-term phenology trends \citep{jeong2013}. Large-scale datasets from China and Europe have already contributed considerably to phenological research \citep{xu2013, olsson2014, basler2016, zhang2017}, and the USA-NPN dataset has the potential to meet these needs for North American plant species and communities. However, the features that allow citizen science projects to collect data at large scales can also introduce spatial biases toward cities and easily-accessible areas, and variation in sampling effort and observer skill \citep{dickinson2010}. With thousands of participants, the potential for variation among observers in their determination of species identification and dating of phenological events is high. While volunteers have been shown to be accurate at distinguishing different leaf and flower stages for plants \citep{fuccillo2015} and can have high agreement on abundance estimates \citep{feldman2018}, contributions to USA-NPN are sometimes made sporadically across seasons, years, and locations. This means that the quantity and quality of data at a specific site will typically be more variable for citizen science efforts than for intensive, long-term studies.

In order to accurately model and forecast phenology, it is important to understand how the strengths and weaknesses of intensive local studies and large-scale citizen science projects influence both our inferences about biological processes driving phenology (e.g. warming requirements for a specific plant) and our ability to predict future phenology events (e.g. forecasting when flowering or leaf out occurs). Here, we fit a suite of plant phenology models for the budburst and first-flowering phenophases of 24 plant species to data from both the USA-NPN and a set of intensive long-term studies from the Long Term Ecological Research (LTER) network. We compare the resulting models based on both inference about models and parameters and predictions for unobserved events. We then use this comparison to assess the best methods for both local- and large-scale phenology modeling and to point the way forward for integrating large-scale and local-scale data to determine the best possible models across scales.

@@

In the LTER datasets observation metrics varied widely due to different protocols. To match the USA-NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods). Three of the LTER datasets (Hubbard Brook, Harvard Forest, and H.J. Andrews) had a sampling frequency of 3-7 days during the growing season. The Jornada dataset had a sampling frequency of 30 days. As with the USA-NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophase combinations which had at least 30 total observations. After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the USA-NPN and LTER datasets to use in the analysis (Table 1 \& S1). Using a 15 day cutoff in the USA-NPN dataset resulted in 35 unique combinations with 23 species.

@@

It is common to fit multiple plant phenology models to find the one that best represents a specific species and phenophase \citep{chuine2013}. For each of the 38 species and phenophase combinations in the five datasets (USA-NPN and four LTER datasets), we fit eight phenology models (Table 2). The \textit{Naive model} uses the mean DOY from prior observations as the estimated DOY. The \textit{Linear model} uses a regression with the mean spring (Jan. 1 - March 31) temperature as the independent variable and DOY as the response variable. For the six remaining models, the general form is based on the idea that a phenological event will occur once sufficient thermal forcing units, $F^{*}$, accumulate from a particular start day of the year ($t_{1}$). Forcing units are a transformation of the daily mean temperature and are calculated differently for each model (Table 2). The start day can either be estimated or fixed. For the \textit{Growing Degree Day (GDD)} model, forcing units are the total degrees above the threshold $T_{base}$ \citep{reaumur1735, wang1960, hunter1992}. The \textit{Fixed GDD model} uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T_{base}$ = 0$^{\circ}$C). The \textit{Alternating model} has a variable number of required forcing units defined as a function of the total number of days below 0$^{\circ}$C since Jan. 1 (number of chill days - $NCD$). The \textit{Uniforc model} is like the \textit{GDD model} but with the forcing units transformed via a sigmoid function \citep{chuine2000}.

@@

As described above, we fit two sets of models for each species and phenophase: one set of models parameterized using only USA-NPN data, and one set parameterized using only LTER data (with the exception of the \textit{M1} and \textit{MSB} models, see above). We performed three primary analyses from these model outputs by comparing: 1) the model parameters, 2) estimates from the models, and 3) out-of-sample errors from each model.

To compare the inferences about process made by the two datasets, we compared the distribution of each parameter between LTER and USA-NPN derived models for each species and phenophase combination. Using the mean value of each bootstrapped parameter, we also calculated the coefficient of determination ($R^2$) between LTER and USA-NPN derived models among the 38 species-phenophases. In three cases where a species phenophase combination occurred in two LTER sites (Budburst for \textit{Acer saccharum}, \textit{Betula alleghaniensis}, and \textit{Fagus grandifolia} in the Harvard and Hubbard Brook datasets) they were compared separately to the USA-NPN data.

Next we compared the estimates of phenological events between models. Models with different parameter values, and even entirely different structures, can produce similar estimates for the date of phenological events \citep{basler2016}. Therefore, to compare the predictions and potential forecasts for models fit to the different datasets, we compared the estimated DOY predicted by the LTER and USA-NPN derived models for all held out observations. For each of the eight models, we calculated the coefficient of determination ($R^2$) between LTER and USA-NPN derived estimates for estimates made at the four LTER sites and across all USA-NPN sites.

Finally, we directly evaluated performance using out-of-sample errors from the four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) USA-NPN derived models predicting LTER observations, C) LTER-derived models predicting USA-NPN observations, and D) USA-NPN derived models predicting USA-NPN observations. Using the RMSE values from held out observations, we compared the performance of LTER and USA-NPN derived models on different data types in two different ways. First, we focused on local-scale prediction by calculating the difference in the RMSE of LTER and USA-NPN derived models solely with LTER observations. Secondly, to focus on large-scale prediction we calculated the difference in RMSE using solely USA-NPN data. These differences were calculated for each of the model types and 38 species-phenophase combinations. Negative values indicate that LTER-derived models perform better, while positive values indicate that the USA-NPN derived model performed better. We used a t-test to test the difference from zero in these values. In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison within each site, not across sites, to focus on local scale prediction when LTER data are available. Absolute RMSE values as well as Pearson correlation coefficients are provided in the supplement for specific species (Fig. S5-S7) and with all observations aggregated together (Table S2).

@@

Throughout the analysis there were no qualitative differences between a 30-day or 15-day threshold between the first 'yes' and most recent 'no' observation in the USA-NPN dataset. Results presented here reflect the 30 day cutoff, see the supplementary figures S2-S4 for matching figures using a 15 day cutoff.

The best matches between parameter estimates based on USA-NPN and LTER data were the \textit{Fixed GDD model} ($R^2$ = 0.49) and the \textit{Linear model} ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$). The parameters for all other models had $R^2$ values \textless 0 indicating that the relationship was worse than no relationship between the parameters (but with matching mean parameter values across the two sets of models) (Fig. 2). The \textit{Naive model} showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTER datasets being mostly in the northern United States compared to the site locations of the USA-NPN dataset (Fig. 2). The large outlier for the \textit{Fixed GDD model} is \textit{Larrea tridentata}; this species' flower phenology is largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}. While the \textit{Fixed GDD} and \textit{Linear} models showed reasonable correspondence between parameter estimates, all parameters for individual species and phenophase combinations had different distributions between USA-NPN and LTER-derived models (Fig. S10-S11).

@@

When comparing the predictive performance using out-of-sample errors, USA-NPN derived models made more accurate predictions for held-out USA-NPN observations, and LTER-derived models performed better on held-out LTER observations ( all \textit{p} < 0.001, Fig. 4). The \textit{Naive} and \textit{Linear} models had the largest differences between the two model sets, while the \textit{Fixed GDD model} had relatively similar errors when evaluated on both USA-NPN and LTER held-out observations. Although the \textit{Fixed GDD model} had the highest agreement in accuracy between USA-NPN and LTER-derived models, it was not the best performing model overall. The \textit{GDD} and \textit{Uniforc} models made the best out of sample predictions, having the lowest RMSE and Pearson coefficient when aggregating all observations together (Table S2). One exception was that the Fixed GDD model had a slightly higher Pearson value when using LTER-derived models to make predictions for USA-NPN data. The best model for specific species and phenophases varied, but was commonly the \textit{Uniforc} and \textit{GDD} models (Figs. S5-S6).

@@

In this study, parameter estimates differed widely within the same phenology model when fit to the two different types of data, except for the simplest process-oriented model: the \textit{Fixed GDD} (Fig. 2). These differences may be caused by a variety of factors that have different implications for interpreting process-oriented models and their parameters. First, the differences could result from limitations in the sampling of the USA-NPN dataset, such as irregular sampling of the same location within or between seasons, leading to less accurate parameter estimates. If this is the case, it would suggest that using LTER data is ideal for making inferences about plant physiology, and that focusing on the \textit{Fixed GDD model} is best for making inferences when USA-NPN data are all that is available. Second, spatial variation (e.g. from local adaptation, acclimation, microclimates, or plant age)  in phenology requirements and drivers could contribute to these differences \citep{diez2012, zhang2017}. Models built using USA-NPN data integrate over that spatial variation, while models built using LTER data only estimate the phenological requirements for a specific site. In this case, USA-NPN data would provide a better estimate of the general phenological requirements of a species, but LTER data would provide a more accurate understanding for a single site. The best solution to this issue would be the development of models that accurately incorporate spatial variation, such as including genetic variation between different populations \citep{chuine2017}, although localized models could also be generated when large-scale predictions are unnecessary. Third, these differences could result from issues with model identifiability: since different parameter values can yield nearly identical estimates of phenological events, parameter estimates can differ between datasets even when the underlying processes generating the data are the same. Information about which of these issues may be causing the differences between datasets can be explored using the analyses in the current study, as will be explained below.

Despite substantial differences in parameter estimates, LTER and USA-NPN derived models produced similar estimates for phenological events in most cases (Fig. 3). This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues. For example, two \textit{GDD} models with parameters of $t_{1}$=1, $F$=10, $T_{base}$=0 and $t_{1}$=5, $F$=5, $T_{base}$=0 produce nearly identical estimates in many scenarios. This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters. In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both USA-NPN and LTER data (Fig. S10-S11). Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Fig. S8-S9). These results are consistent with research showing that models failed to estimate the starting day of warming accumulation solely from budbreak time-series, thus producing parameter estimates that were not biologically realistic \citep{chuine2016}. \cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting. Here, our simplest model, the \textit{Fixed GDD model} which uses only a warming component, had the highest correlation among parameters between LTER and USA-NPN datasets. In combination with the aforementioned studies, our results indicate that caution is warranted in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.

@@

In addition to exploring differences between phenology datasets, our analyses provide guidance on which models to use when making predictions at a local scale using models built from large-scale data, or vice versa. Among the eight models tested, the \textit{Uniforc} and \textit{GDD} models performed the best overall in the cross dataset comparison in terms of Pearson correlation and RMSE (Fig. S5-S6, Table S2). The \textit{GDD} model has one less parameter than the \textit{Uniforc} model, thus the \textit{GDD} model is a suitable choice for making predictions when there is little to no information at the location of interest (e.g. making phenology forecasts at a new location distant from any observed data). This guidance can vary between species though and model testing should still be performed when suitable data is available.

In conclusion, our results suggest that both LTER and USA-NPN data provide valuable information on plant phenology. Models built using both data sources yield effective predictions for phenological events, but parameter estimates from the two data sources differ and models from each source best predict that data source's phenology events. The primary difference in the datasets is spatial scale, but due to trade-offs in data collection efforts, the larger scale USA-NPN data have shorter time-series, less site fidelity and other differences from the intensively collected LTER data (Table 3). These differences can be strengths or potential limitations. Observers sampling opportunistically allows the USA-NPN dataset to have a large spatial scale, but also leads to low site fidelity, which limits the ability to measure long-term trends at local scales \citep{gerst2016}. Tracking long-term trends is the major strength of LTER data, but having a relatively small species pool limits their use in species-level predictive modeling. Due to these differences, the best data source for making predictions depends on the scale at which the predictions are being made. Identifying the most effective data sources for different types and scales of analysis is a useful first step, but the ultimate solution to working with diverse data types is to focus on integrating all types of data into analyses and forecasts \citep{hanks2018, melaas2016}. Our results suggest that methods that can learn from the intensive information available in LTER data in regions where they are available, and simultaneously use large-scale data to capture spatial variation in phenological requirements will help improve our ability to understand and predict phenology. Data integration efforts should also leverage data from remote sensing sources such as the PHENOCAM network or satellite imagery, which have both a large spatial extent and high temporal resolution \citep{peng2017,richardson2018a,richardson2018b}. Data integration provides the potential to use data from many sources to produce the best opportunity for accurate inference about, and forecasting of, the timing of biological events.

@@

This research was supported by the Gordon and Betty Moore Foundation’s Data-Driven Discovery Initiative through Grant GBMF4563 to E.P. White. J.M. Meiners was supported by the University of Florida Biodiversity Institute Graduate Research Fellowship. We thank the developers and providers of the data that made this research possible including: the USA National Phenology Network and the many participants who contribute to its Nature’s Notebook program; the H.J. Andrews Experimental Forest research program, funded by the National Science Foundation's (NSF) LTER Program (DEB-1440409), US Forest Service Pacific Northwest Research Station, and Oregon State University; the Jornada Basin LTER project (NSF Grant DEB-1235828); the Hubbard Brook Experimental Forest, which is operated and maintained by the USDA Forest Service, Northern Research Station, Newtown Square, PA.; the Harvard Forest LTER; and the PRISM Climate Group at Oregon State University.
